%! Author = grays
%! Date = 3/30/2026

% Preamble
\documentclass[11pt]{article}

% Packages
\usepackage{amsmath}
\usepackage{caption}
\usepackage{float}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{mhchem}
\usepackage{siunitx}
\usepackage{xcolor}
\usepackage[backend=biber,style=ieee]{biblatex}
\usepackage[colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue]{hyperref}
% Listings settings
\lstset{
    basicstyle=\ttfamily\footnotesize,
    keywordstyle=\color{blue},
    commentstyle=\color{green!50!black},
    stringstyle=\color{orange},
    numbers=left,
    numberstyle=\tiny\color{gray},
    stepnumber=1,
    numbersep=5pt,
    backgroundcolor=\color{gray!10},
    showstringspaces=false,
    breaklines=true,
    frame=single,
    rulecolor=\color{gray!50},
    title=\lstname
}
\sisetup{separate-uncertainty=true}
\DeclareSIUnit\angstrom{\text {Å}}

\addbibresource{references.bib}

\title{Spectroscopy}
\date{April 3, 2026}
\author{
    Patrick Geraghty (300349204) \\
    Evan Champagne (300443446)
}

% Document
\begin{document}
\maketitle

    \section{Introduction}\label{sec:introduction}

        Spectroscopy deals with the analysis of materials or gases by
        measurements of their absorption, reflection, or emission spectra.
        In this experiment, a grating spectrometer apparatus---consisting
        of a light source, collimator, grating, and telescope---is used to
        study the emission spectra of low-pressure gases in which atoms are
        excited by an electrical discharge.

        The primary purpose of this experiment is to explore these atomic and
        molecular emissions.
        Firstly, the grating spectrometer is calibrated using the well-known
        emission doublet of sodium (\ce{Na}), which has two closely spaced
        lines at \SI{5889.95}{\angstrom} and \SI{5895.92}{\angstrom}.
        The fundamental principle governing this is the grating equation
        \begin{equation}
            m \lambda = d (\sin \theta)
            \label{eq:grating}
        \end{equation}
        where \( d \) represents the grating constant, \( m \) is the order of
        diffraction, and \( \theta \) is the diffraction angle for a specific
        wavelength \( \lambda \).
        Next, utilizing the calibrated spectrometer, the emission lines of
        a hydrogen source are measured, alongside those of helium, nitrogen,
        and an unknown gas.
        These measurements are then compared against established spectral
        data, and the unknown gas is identified based on its emission
        characteristics.
        Lastly, the hydrogen spectrum is analyzed to determine the Rydberg
        constant, \( R \).
        This is calculated using the quantized energies of an electron in a
        Coulomb potential, given by
        \begin{equation}
            \frac{1}{\lambda} = R \left( \frac{1}{4} - {1}{n^2} \right)
        \end{equation}
        where \( n \) is the principal quantum number of the Balmer series
        transitions for each observed spectral line.

    \section{Methods}\label{sec:methods}

        \subsection*{Experimental Setup}
        The experimental setup consists of a collimator, a diffraction
        grating (600 lines/mm), and a telescope mounted on a rotating stage.
        A discharge tube is placed at the focal point of the collimator to
        serve as the light source.
        The light passes through the collimator to create a parallel beam,
        strikes the diffraction grating, and is then observed through the
        rotating telescope, allowing for the measurement of the angles
        corresponding to different spectral lines.

        \subsection*{Data Collection}
        The spectrometer is first calibrated using a sodium lamp, where
        a reference initial angle is recorded with the telescope approximately
        normal to the grating.
        The angles for the first-order diffraction lines are measured, and the
        corresponding wavelengths are calculated using the grating equation.
        The calculated wavelengths are then compared to the known values to
        determine whether the starting angle is accurate or if an adjustment
        is necessary.

        After calibration, the sodium lamp is replaced with discharge tubes
        containing hydrogen, helium, nitrogen, and an unknown gas (referred to
        as Unknown E).
        For each gas, the telescope is swept across the viewing field to
        locate the visible emission lines, and the corresponding angles are
        recorded.

        \subsection*{Data Analysis}
        The raw angular data is processed using a Python script that automates
        the calibration, wavelength calculation, and determination of the
        Rydberg constant for hydrogen.

    \section{Raw Data}\label{sec:data}

    \begin{table}[h!]
        \centering
        \caption{Spectral Emission Observations for Various Gases}
        \label{tab:spectral_observations}
        \begin{tabular}{lccc}
        \hline
        \textbf{Gas} & \textbf{Base Angle ($^\circ$)} & \textbf{Spectral Line Color} & \textbf{Observation Angle ($^\circ$)} \\ \hline
        Sodium & 295.0 & Doublet (Na) & 313.66 \\ \hline
        Hydrogen & 295.0 & Violet & 293.0 \\
         & & Blue & 277.5 \\
         & & Cyan & 276.6 \\
         & & Red & 269.0 \\ \hline
        Helium & 295.0 & Purple & 277.0 \\
         & & Cyan & 276.0 \\
         & & Green & 275.3 \\
         & & Yellow & 275.0 \\
         & & Orange & 271.6 \\
         & & Red & 268.6 \\ \hline
        Nitrogen & 295.0 & Distinct & 293.0 \\
         & & Blue lines (max) & 278.6 \\
         & & Blue lines (min) & 276.0 \\ \hline
        Unknown E & 295.0 & Blue 1 \& 2 & 277.6 \\
         & & Blue 3 & 277.3 \\
         & & Blue 4 \& 5 & 277.0 \\
         & & Green & 273.0 \\
         & & Orange & 271.6 \\
         & & Red & 269.0 \\ \hline
        \end{tabular}
    \end{table}

    \section{Analysis and Results}\label{sec:analysis}

    \section{Discussion}\label{sec:discussion}
        Comparing the observed spectral lines to known emission
        spectra~\cite{nist_helium}

    \section{Conclusion}\label{sec:conclusion}

    \printbibliography

\end{document}